---
title: "3k running assignment"
author: "your name, id="
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
 word_document: default
 # html_document: default
 # if you don't have Microsoft Word, you can output an html document and save it as a pdf for submission.
 # just uncomment (remove the #) from the "# html_document: default" line and add a # in front of the "word_document: default" line
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Inference and linear regression for 3 km Running Times based on laboratory testing

### Study Description:

Sixteen well-trained male middle and long distance runners performed independently a 3 km time trial and a number of running tests in the laboratory.  

One of the tests was running velocity at a blood lactate concentration of 4 mmol/L (`v4mM`), measured in km/hr.  For background knowledge to understand something about the study, it requires considerable physical effort to achieve a blood lactate concentration of 4 mmol/L in well-trained athletes (if you want to know more about blood lactate and athletic performance here is a reference https://www.trainingpeaks.com/blog/what-is-lactate-and-lactate-threshold/).  

Other tests were running velocity at Lactate Threshold (`vTlac`), and at VO2 max (`V02Max`) both measured in km/hr.  In addition, a measure of running economy, oxygen uptake (measured in ml/kg/min) at running speeds of 14.5km/hr and 16.1 km/hr were measured and the variable names are `Rel.14.5` and `Rel.16.1`.  

All the laboratory testing took place on a motorised treadmill, and distance running performance was determined by 3 km time trials on an indoor 200m track, 3k time in minutes was recorded as `Running.Time`.

Based on data on other populations, the sports scientist doing the study wants to know what evidence the sample provides that the mean 3k running time in this population is different to 10 minutes 12 seconds (10.2 minutes), and also wants to estimate a plausible interval for the population mean 3k running time.  The sports scientist also wants to estimate the 3k time below which the fastest 10% of the population times lie.  Finally, the sports scientist wants to investigate if it is possible to predict 3k running time from running velocity at a blood lactate concentration of 4 mmol/L.

### Aims: 

To investigate the following:

(1) Is the population mean 3k running time different to 10 minutes 12 seconds (ie as decimal 10.2 minutes)?  What is a plausible range in which the population mean 3k time is likely to lie?  And what time are we confident that the top 10% of the population 3k times will be faster than?

(2) Can we use linear regression to predict 3 km running time in minutes (`Running.Time`) from running speed at lactate 4 mmol/L in km per hour (`v4mM`)?

It is decided in advance that the acceptable significance level for hypothesis testing is $\alpha$ = 0.05.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# load required libraries
library(tidyverse)
library(infer)
library(tolerance)
```

### Read the data and see a few rows

```{r}
running = read.csv("3krunning.csv", header = TRUE)
head(running)
```

### Summary Statistics

Shown below are summary statistics for some of the columns in the dataset. 

```{r}
running %>% summarise(
  count = n(),
  meantime = mean(Running.Time),
  sdtime = sd(Running.Time),
  meanv4mM = mean(v4mM),
  sdv4mM = sd(v4mM),
  meanRel14.5 = mean(Rel.14.5),
  sdRel14.5 = sd(Rel.14.5),
  
       )

```


## Hypothesis test and confidence interval for mean 3k time

#### Task 1: Write the null and alternative hypotheses and significance level of the test (text, 1 sentence)
i.e. the hypothesis test for Aim (1), first sentence.
 
**5 marks**

#### Task 2): Appropriate statistical test and assumptions

(a) What statistical test is appropriate to test the Task 1 hypotheses?  State the assumptions needed for the test to be valid, and how they could be justified in this study. (1-2 sentences).   

(b) Comment on the output of the R code below in relation to an assumption for the test (1 sentence)

```{r}

running %>% ggplot(aes(x =Running.Time, y = "" )) + geom_boxplot() + labs(title = "Boxplot of 3k running times in sample") 

```

**10 marks**

#### Task 3: Carry out the hypothesis test and estimate a 95% confidence interval for the true mean 3k time

The code below does a two-sided t test for the null hypothesis that the true mean `Rel.14.5` is equal to 50. Adapt the code below to carry out the hypothesis test from Task 1.


```{r}
running %>% dplyr::select(Rel.14.5) %>% t.test(mu = 50)
```
**5 marks**

#### Task 4: Interpret (a) the results of the hypothesis test (b) the 95% confidence interval from Task 3 

Note: you need to interpret the results after you have adapted the code in Task 3.  You will not get marks for this question if you intepret the hypothesis test for the `Rel.14.5` variable.
(1-2 sentences)

**5 marks**

#### Task 5: A bootstrap 95% confidence interval for the mean 3k time 

The code below estimates a 90% confidence interval for the mean 3k time using the bootstrap method.
Change the number of bootstrap replications to 1000, and change the confidence level to 95%. 


```{r}
set.seed(123456789) # this is for replicable simulations - do not change this
boot <- running %>% 
          specify(response = Running.Time) %>%
          generate(reps = 100, type = 'bootstrap') %>%
          calculate(stat = 'mean')

### Produce the 90% bootstrap confidence interval
### using the generated bootstrap distribution of the sample statistic
get_ci(boot, level = 0.90)

### plot the histogram of bootstrap statistics
### and shading the bootstrap confidence interval. 
boot %>% 
  visualize(method = "simulation") +
  shade_confidence_interval(get_ci(boot))+
  xlab("Bootstrap Mean")+
  ylab("Frequency")


```

**5 marks**

#### Task 6: Compare the 95% confidence interval using the t distribution with the bootstrap 95% CI 

(1-2 sentences)


**5 marks**

#### Task 7: A tolerance interval for the range in which we are 95% confident 90% of the population of 3k times will lie.  

(a) The code below estimates a tolerance interval for the range in which we are 95% confident 95% of the population of 3k times will lie.  Adapt the code to estimate  a tolerance interval for the range in which we are 95% confident 90% of the population of 3k times will lie.

(b) The men's world record for the 3k is 7 mins 20.67 seconds by Daniel Komen of Kenya in 1996.  Using your adapted tolerance interval, state the running time you are 95% confident that the fastest 10% of our study population will achieve. (1 sentence) 

```{r}
normtol.int(running$Running.Time, alpha = 0.05, P = 0.95, side = 2)
```


**5 marks**

## Linear Regression


#### Task 8: Labelled scatterplot with smooth of `Running.Time` versus `v4mM`

(a) The code below makes a scatterplot with smooth of `Running.Time` versus `Rel.14.5`.  Adapt the code to make a scatterplot with smooth of `Running.Time` versus `v4mM`.  Remember to update the title, axis labels, and units. 

(b) What does the smoother suggest regarding the suitability of a simple linear regression model for the relationship between `Running.Time` and `v4mM`? (1 sentence)

```{r, warning = FALSE, message = FALSE}
running %>% ggplot(aes(x = Rel.14.5, y = Running.Time)) + geom_point() + geom_smooth() + labs(title = "Scatterplot with smooth for 3k time versus running economy at 14 kph", x = "VO2 at 14.1 kph (ml/kg/min)", y = "3k time (minutes)")
```


**5 marks**

#### Task 9: Correlation coefficient between `v4mM` and `Running.Time` 

(a) The code below calculates the sample correlation coefficient between `Running.Time` and `Rel.14.5`.  Change to code to calculate the sample correlation coefficient between `Running.Time` and `v4mM`.

```{r}
running %>% dplyr::select (Running.Time, Rel.14.5) %>% cor()
```


(b) Interpret the relationship between `Running.Time` and `v4mM`, referring to the relevant scatterplot and correlation coefficient.


**10 marks**
  
#### Task 10: Line of best fit for the linear relationship between 3 km running time and v4mM in the sample.  

(a) The code below calculates the simple linear regression for `Running.Time` versus `Rel.14.5`.  Change the code to produce the simple linear regression for `Running.Time` versus `v4mM`.

```{r}
lm1<-lm(Running.Time ~ Rel.14.5, data = running)
lm1
```

**5 marks**

#### Task 11: Write down the equation of the line of best fit from for the simple linear regression of `Running.Time` on `v4mM` and provide an interpretation of the slope and intercept. Does the intercept have a physically meaningful interpretation? 

(2-3 sentences)

**10 marks**

#### Task 12: Predict the running time (i.e. `Running.Time`) when running speed at blood lactate concentration 4 mmol/litre (i.e. `v4mM`) is 18.9 km per hour.

The code below predicts `Running.Time` when `Rel.14.5` = 48 ml/kg/min from the linear regression model of `Running.Time` on `Rel.14.5`.  It also shows the first few rows of the dataset.

(a) Adapt the code to predict `Running.Time` when `v4mM`=18.9 km/hr from a linear regression of `Running.Time` on `v4mM`.

(b) Why is the predicted running time at `v4mM`=18.9 different to 8.82, the observed running time when `v4mM` is 18.9 km/hr? - see observation 4 in the dataset.  What is the value of the residual for this observation?  

(1-2 sentences)

```{r}
lm1<-lm(Running.Time ~ Rel.14.5, data = running)
newdata <-data.frame(Rel.14.5 = 48)
predict(lm1, newdata = newdata)
head(running)
```


**10 marks**

#### Task 13: Predict the running time (i.e. `Running.Time`) when  `v4mM` is 2.6 km per hour. Explain if you have any concern related to this prediction.  

(a) Write code in the chunk below to predict `Running.Time` when `v4mM`= 2.6 km/hr.  (Hint: copy and adapt the relevant code from the chunk above)

(b) Do you have any concern about this prediction, and if so why? (1 sentence)

```{r}

```


**10 marks**

## Overall Conclusion and presentation 

#### Task 14: State your overall conclusions in relation to the two main aims of the study (3-4 sentences for each)

Aim (1) From the hypothesis test, confidence intervals for the true mean, and tolerance interval.

**5 marks**

Aim (2) From fitting a linear model for the relationship between 3k running time and the running speed at blood lactate concentration 4 mmol/litre.   

**5 marks**

#### Task 15: Knit your .Rmd file to Word and submit the resulting Word document (alternatively knit to html, save the html as a pdf and submit the pdf).  Make sure you have edited the .Rmd to include your own name and student ID.  

If you submit an .Rmd file you will **lose 20 marks**, and if you submit a file missing your name and/or ID you will **lose 10 marks**.

