---
title: "Assignment 2"
author: "myname, id=11111111"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Simple example

I have created a vector of data called `data1` and have calculated some numerical and graphical summaries.

These are the data:

```{r create data}

data1 <- c(10, 23, 14, 12, 34, 26, 28)

```

The mean is

```{r calculate mean}
mean(data1)

```

The summary statistics (minimum, maximum, $Q_1$, median, and $Q_3$) are:

```{r summary stats}

```


The five number summary which uses Tukey's method to estimate the lower and upper quartiles ($Q_1$ and $Q_3$) is given below. Sometimes there are small differences between the quartiles given by the `summary` function and quartiles given using Tukey's method.





A boxplot of the data below uses Tukey's method. I would describe the shape of the distribution using the boxplot as ...

```{r}

boxplot(data1)

```


A histogram with R's default settings is shown below. I would describe the shape of the distribution using the histogram as ...      





A histogram with about ten bins is also shown below.


